---
title: "Spatial statistics in R"
author: "Cornelius Senf"
date: "cornelius.senf@geo.hu-berlin.de"
output:
  ioslides_presentation:
    smaller: yes
    widescreen: yes
  beamer_presentation: default
  slidy_presentation: default
subtitle: Quantitative Methods - Lab Session 16
header-includes: \usepackage{amsmath}
---

<style type="text/css">
slide.backdrop {
  background: none !important;
  background-color: white !important;
}
</style>

```{r echo=FALSE, hide=TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
library(tidyverse)
library(sp)
library(mapview)
```

## Today

- Detecting spatial auto-correlation in regression models
- Spatial lag models
- Spatial error models

## Spatial residual analysis

Let's build a simple linear regression model predicting the zinc concentration based on the elevation and squarroot-transformed distance:

```{r}
data("meuse")

fit <- lm(log(zinc) ~ elev + sqrt(dist), data = meuse)
```

## Spatial residual analysis

```{r}
summary(fit)
```

## Spatial residual analysis

```{r}
resid <- data.frame(resid = residuals(fit),
                    fitted = fitted(fit))

p1 <- ggplot(resid, aes(x = resid)) + geom_histogram()

p2 <- ggplot(resid, aes(x = 1:nrow(meuse), y = resid)) + geom_point()

p3 <- ggplot(resid, aes(x = fitted, y = resid)) + geom_point()
```

## Spatial residual analysis

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center'}
print(p1)
```

## Spatial residual analysis

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=7, fig.align='center'}
print(p2)
```

## Spatial residual analysis

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=7, fig.align='center'}
print(p3)
```

## Spatial residual analysis

The residuals look fine. There's no serial correlation or heteroskedasticity. However, are the residuals spatially independent?

```{r}
meuse$residuals <- fit$residuals

p <- ggplot(meuse, aes(x = x, y = y, col = residuals, size = residuals)) +
  geom_point() +
  scale_color_gradient2()
```

## Spatial residual analysis

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=6, fig.align='center'}
print(p)
```

## Spatial residual analysis

Testing for spatial autocorrelation:

```{r, warning=FALSE, message=FALSE}
library(spdep)

dnn <- dnearneigh(as.matrix(meuse[,c("x", "y")]), d1 = 0, d2 = 200)
neighw <- nb2listw(dnn, style = "W", zero.policy = TRUE)
moran <- moran.mc(fit$residuals, neighw, nsim = 999, zero.policy = TRUE)

moran
```

## Spatial residual analysis

Sensitivity analysis:

```{r}
moran_I <- c()

for (d in seq(50, 2000, 50)) {
  dnn <- dnearneigh(as.matrix(meuse[,c("x", "y")]), d1 = 0, d2 = d)
  neighw <- nb2listw(dnn, style = "W", zero.policy = TRUE)
  moran <- moran.mc(fit$residuals, neighw, nsim = 999, zero.policy = TRUE)
  moran_I <- c(moran_I, moran$statistic)
}

moran_I <- data.frame(moran = moran_I, 
                      distance = seq(50, 2000, 50))

p <- ggplot(moran_I, aes(x = distance, y = moran)) + 
  geom_point() +
  geom_line()
```

## Spatial residual analysis

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=7, fig.align='center'}
print(p)
```

## Spatial regression models

We can use spatial lag models (or spatial simultaneous autoregressive lag model) to encounter spatial auto-correlation in the residuals. Spatial lag models incorporate an additional parameter of lagged y-values:

$y = \rho * W * y + X\beta + e$ with $e \sim N(0, \sigma)$

Where $\rho$ estimates the influence of neighboring values of y and $W$ defines a weighted neighborhood matrix.

## Spatial regression models

Building a spatial lag model in R is easy using the `lagsarlm()` function in the `spdep` package:

```{r}
dnn <- dnearneigh(as.matrix(meuse[,c("x", "y")]), d1 = 0, d2 = 400)
neighw <- nb2listw(dnn, style = "W", zero.policy = TRUE)

fit_sp1 <- lagsarlm(log(zinc) ~ elev + sqrt(dist), data = meuse, listw = neighw, zero.policy = TRUE)
```

##

```{r, echo=FALSE}
summary(fit_sp1)
```

## Spatial regression models

```{r}
meuse$residuals_sp1 <- fit_sp1$residuals

p <- ggplot(meuse, aes(x = x, y = y, col = residuals_sp1, size = residuals)) +
  geom_point() +
  scale_color_gradient2()
```

## Spatial regression models

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=6, fig.align='center'}
print(p)
```

## Spatial regression models

The spatial lag model didn't account for our spatially auto-correlated residuals. Alternatively, we can use spatial error model, which incorporate an additional parameter of lagged residual values:

$y = X\beta + u$ with $U = \lambda * W * u + e$ and $e ~ N(0, \sigma)$

## Spatial regression models

Building a spatial error model in R is easy using the `errorsarlm()` function in the `spdep` package:

```{r}
dnn <- dnearneigh(as.matrix(meuse[,c("x", "y")]), d1 = 0, d2 = 400)
neighw <- nb2listw(dnn, style = "W", zero.policy = TRUE)

fit_sp2 <- errorsarlm(log(zinc) ~ elev + sqrt(dist), data = meuse, listw = neighw, zero.policy = TRUE)
```

##

```{r, echo = FALSE}
summary(fit_sp2)
```

## Spatial regression models

```{r}
meuse$residuals_sp2 <- fit_sp2$residuals

p <- ggplot(meuse, aes(x = x, y = y, col = residuals_sp2, size = residuals)) +
  geom_point() +
  scale_color_gradient2()
```

## Spatial regression models

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=6, fig.align='center'}
print(p)
```

## Spatial regression models

```{r}
moran_I <- c()
moran_I_sp1 <- c()
moran_I_sp2 <- c()

for (d in seq(50, 2000, 50)) {
  dnn <- dnearneigh(as.matrix(meuse[,c("x", "y")]), d1 = 0, d2 = d)
  neighw <- nb2listw(dnn, style = "W", zero.policy = TRUE)
  
  moran <- moran.mc(fit$residuals, neighw, nsim = 999, zero.policy = TRUE)
  moran_I <- c(moran_I, moran$statistic)
  
  moran_sp1 <- moran.mc(fit_sp1$residuals, neighw, nsim = 999, zero.policy = TRUE)
  moran_I_sp1 <- c(moran_I_sp1, moran_sp1$statistic)
  
  moran_sp2 <- moran.mc(fit_sp2$residuals, neighw, nsim = 999, zero.policy = TRUE)
  moran_I_sp2 <- c(moran_I_sp2, moran_sp2$statistic)
}
```

## Spatial regression models

```{r}
moran_I <- data.frame(moran = moran_I, 
                      moran_sp1 = moran_I_sp1, 
                      moran_sp2 = moran_I_sp2, 
                      distance = seq(50, 2000, 50))

p <- ggplot(moran_I) + 
  geom_point(aes(x = distance, y = moran)) +
  geom_line(aes(x = distance, y = moran)) + 
  geom_point(aes(x = distance, y = moran_sp1), col = "red") +
  geom_line(aes(x = distance, y = moran_sp1), col = "red") + 
  geom_point(aes(x = distance, y = moran_sp2), col = "blue") +
  geom_line(aes(x = distance, y = moran_sp2), col = "blue")
```

## Spatial regression models

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=7, fig.align='center'}
print(p)
```

## Spatial regression models

```{r}
AIC(fit, fit_sp1, fit_sp2)
```
