---
title: "Assignment 6"
author: ""
date: ""
output: html_document
---

# Does the diving time of cormorants differ with season and subspecies? 

A team of researchers has tracked the seasonal diving time of two sub-species of cormorants (Phalocrocorax carbo). The sub-species are Phalocrocorax carbo carbo and Phalocrocorax carbo sinensis. You're asked to analyze whether there are significant differences in diving time between the two sub-species, as well as between seasons.

## 1) Import and visualize the data (2 Points)

Import the cormorant dataset. The dataset contains diving times (seconds) for the two subspecies. Subspecies are coded as C and S. Create a boxplot of the diving times by season and subspecies. The seasons are coded numerically as follows: 1 (Spring), 2 (Summer), 3 (Autumn), and 4 (Winter). Note, you must convert the variable season into a factor variable.

```{r}

```

## 2) Test for variance homogeneity (2 Point)

```{r}

```

Describe your findings in one sentence:

## 3) Test the signficance of the effects of season and sub-species on diving time (2 Point)

```{r}

```

Describe your findings in one sentence:

## 4) Check the model assumptions (homoskedasticity and normality of residuals) (2 Points)

```{r}

```

Describe your findings in 1-2 sentences.

# 5) Identify which seasons differ with respect to diving time (2 Points)

Conduct a multiple-comparison test to assess which seasons are significantly different from each other. Describe your findings in 1-2 sentences.

```{r}

```

