---
title: "Spatial statistics in R"
author: "Cornelius Senf"
date: "cornelius.senf@geo.hu-berlin.de"
output:
  ioslides_presentation:
    smaller: yes
    widescreen: yes
  beamer_presentation: default
  slidy_presentation: default
subtitle: Quantitative Methods - Lab Session 15
header-includes: \usepackage{amsmath}
---

<style type="text/css">
slide.backdrop {
  background: none !important;
  background-color: white !important;
}
</style>

```{r echo=FALSE, hide=TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
library(tidyverse)
library(sp)
library(mapview)
```

## Today

- Semivariogram
- Kriging
- Neighborhood definitions
- Moran's I

## Data

We're today usinf the `meuse` dataset, which is provided by `sp` package. It comprises of measurements of four heavy metalsin the top soil in a flood plain along the river Meuse, along with a handful of covariates.

```{r}
data("meuse")

head(meuse)
```

## Data

Converting it into a spatial points object:

```{r, eval=FALSE, warning=FALSE, message=FALSE}
library(sp)

coordinates(meuse) <- ~x+y
proj4string(meuse) <- CRS("+init=epsg:28992")

mapview(meuse)
```

## Data

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(sp)
library(mapview)

coordinates(meuse) <- ~x+y
proj4string(meuse) <- CRS("+init=epsg:28992")

mapview(meuse)
```

# Semivariogram

## Semivariogram

Semivariograms show the semivariance (variance of a variable when compared at distant locations). We can estimate an empirical semivarigrams using the `gstat` package:

```{r}
library(gstat)

g <- gstat(formula = log(zinc) ~ 1, data = meuse)
var <- variogram(g)

head(var)
```

## Semivariogram

Plotting the semivariogram:

```{r}
p <- ggplot(var, aes(x = dist, y = gamma, size = np)) + 
  geom_point()
```

## Semivariogram

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=7, fig.align='center'}
print(p)
```

## Semivariogram

We can use the empirical semivariogram to fit a theoretical function describing the semivariance at each distance. There are several models (exponential, spherical, gaussian, etc.) and we can utilize the `fit.variogram` function to derive the best-fitting model:

```{r}
var_fit <- fit.variogram(var, model = vgm(c("Exp", "Sph", "Gau")))

var_fit
```

See https://cran.r-project.org/web/packages/gstat/vignettes/gstat.pdf for further information on semi-variograms!

See `vgm()` for all available models.

## Semivariogram

Adding the theoretical semivariogram to the plot:

```{r}
vari_fit_line <- variogramLine(var_fit, maxdist = max(var$dist))

p <- ggplot(var, aes(x = dist, y = gamma)) + 
  geom_point() +
  geom_line(data = vari_fit_line, aes(x = dist, y = gamma))
```

## Semivariogram

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=7, fig.align='center'}
print(p)
```

# Kriging

## Kriging

Kriging utilizes the theoretical semivariogram to interpolate values at any location based on distant-variance relationship. Before we can perform Kriging, we need to create a result-layer of points we want to interpolate:

```{r}
xnew <- seq(meuse@bbox[1, 1], meuse@bbox[1, 2], length.out = 50)
ynew <- seq(meuse@bbox[2, 1], meuse@bbox[2, 2], length.out = 50)

gridnew <- expand.grid(xnew, ynew)

head(gridnew)
```

## Kriging

```{r, eval=FALSE, warning=FALSE, message=FALSE}
names(gridnew) <- c("x", "y")
coordinates(gridnew) <- ~x+y
proj4string(gridnew) <- CRS("+init=epsg:28992")
gridded(gridnew) <- TRUE

mapview(gridnew)
```

## Kriging

```{r, echo=FALSE, warning=FALSE, message=FALSE}
names(gridnew) <- c("x", "y")
coordinates(gridnew) <- ~x+y
proj4string(gridnew) <- CRS("+init=epsg:28992")
gridded(gridnew) <- TRUE

mapview(gridnew)
```

## Kriging

```{r, message=FALSE, warning=FALSE}
kriging <- krige(log(zinc) ~ 1, meuse, gridnew, model = var_fit)
kriging <- as.data.frame(kriging)

p <- ggplot(kriging, aes(x = x, y = y)) + 
  geom_point(aes(col = var1.pred)) +
  scale_color_gradient(low = "yellow", high = "red") +
  geom_point(data = as.data.frame(meuse@coords), aes(x = x, y = y)) +
  coord_equal()
```

## Kriging

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=7, fig.align='center'}
print(p)
```

## Kriging

We can smooth the figure by rasterizing the data:

```{r}
p <- ggplot(kriging, aes(x = x, y = y)) + 
  geom_tile(aes(fill = var1.pred)) +
  scale_fill_gradient(low = "yellow", high = "red") +
  geom_point(data = as.data.frame(meuse@coords), aes(x = x, y = y)) +
  coord_equal()
```

## Kriging

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=7, fig.align='center'}
print(p)
```

# Neighborhood definitions

## Neighborhood definitions

Using points we can define spatial neighborhoods based on their k nearest neighbors:

```{r, warning=FALSE, message=FALSE}
library(spdep)

knn <- knearneigh(meuse@coords, k = 1)
knn <- knn2nb(knn)
knn
```

## Neighborhood definitions

Maximum k of 1, 2, 3, 4:

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, fig.width=10, fig.align='center'}
knn2 <- knn2nb(knearneigh(meuse@coords, k = 2))
knn3 <- knn2nb(knearneigh(meuse@coords, k = 3))
knn4 <- knn2nb(knearneigh(meuse@coords, k = 4))

par(mfrow = c(1, 4))
plot(knn, meuse@coords, col = "red")
plot(knn2, meuse@coords, col = "red")
plot(knn3, meuse@coords, col = "red")
plot(knn4, meuse@coords, col = "red")
```

## Neighborhood definitions

Similarely, we can define neighborhoods based on a radial distance:

```{r}
dnn <- dnearneigh(meuse, d1 = 0, d2 = 200)
dnn
```

## Neighborhood definitions

Maximum distances of 200, 300, 400 and 500:

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, fig.width=10, fig.align='center'}
dnn2 <- dnearneigh(meuse, d1 = 0, d2 = 300)
dnn3 <- dnearneigh(meuse, d1 = 0, d2 = 400)
dnn4 <- dnearneigh(meuse, d1 = 0, d2 = 500)

par(mfrow = c(1, 4))
plot(dnn, meuse@coords, col = "red")
plot(dnn2, meuse@coords, col = "red")
plot(dnn3, meuse@coords, col = "red")
plot(dnn4, meuse@coords, col = "red")
```

## Neighborhood definitions

Similarely, we can calculate neighborhoods using polygons. Let's first load and plot a polygon of administrative units in Germany:

```{r, warning=FALSE, message=FALSE}
shp <- raster::shapefile("data/poverty_vietnam.shp")

#mapview(shp)
```

The data are similar to: https://www.ecologyandsociety.org/vol13/iss2/art24/

## Neighborhood definitions

Neighborhoods with polygon are defined by touching edges and/or nodes:

```{r}
nb <- poly2nb(shp, queen = TRUE)
nb
```

## Neighborhood definitions

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, fig.width=4, fig.align='center'}
plot(shp)
plot(nb, coordinates(shp), add = TRUE, col = "red")
```

## Neighborhood definitions

For more information on neighborhood definitions, see: https://cran.r-project.org/web/packages/spdep/vignettes/nb.pdf

# Moran's I

## Moran's I

For calculating Moran's I, we first need to define a weighted neighborhood file:

```{r}
neighw <- nb2listw(nb, style = "W", zero.policy = TRUE)
```

The `zero.policy` arguments allows for units without neighbors. See `?nb2listw` for weighting styles.

## Moran's I

```{r, fig.show='hide'}
moransI <- moran(shp$FGT0, neighw, n = length(nb), S0 = Szero(neighw), zero.policy = TRUE)

moran.plot(shp$FGT0, neighw, zero.policy = TRUE)
```

## Moran's I

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center'}
moran.plot(shp$FGT0, neighw, zero.policy = TRUE)
```

## Moran's I

```{r}
moran.mc(shp$FGT0, neighw, nsim = 999, zero.policy = TRUE)
```

## Moran's I

```{r}
localmoran <- localmoran(shp$FGT0, neighw, zero.policy = TRUE)

head(localmoran)
```

## Moran's I

```{r}
shp@data$id <- rownames(shp@data)
shp@data$localmoran <- localmoran[, 1]

shp_df <- dplyr::left_join(fortify(shp, region = "id"), # convert to a data.frame
                           shp@data,
                           by = "id")

p <- ggplot(shp_df, aes(x = long, y = lat, group = group, fill = localmoran)) + 
  geom_polygon() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  coord_equal()
```

## Moran's I

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, fig.width=4, fig.align='center'}
print(p)
```