---
title: "Introduction to R"
author: "Dirk Pflugmacher"
date: "dirk.pflugmacher@geo.hu-berlin.de"
output:
  ioslides_presentation:
    smaller: yes
    widescreen: yes
  beamer_presentation: default
  slidy_presentation: default
subtitle: Quantitative Methods - Lab Session 03
header-includes: \usepackage{amsmath}
---

# Writing R Programs

## Flow control

- Until now, we have seen components of the R language that cause evaluation of single expressions.

```{r}
i <- sqrt(9)
```

- Remember, expression can consitute a mathematical operation, a logical condition, or any function for that matter.
- However, R is a true programming language that allows **conditional execution** and **looping constructs** as well. 

## Flow control: *conditions*

In previous sessions we learned about logical vectors and operators and that they can be used to perform different operations on different parts of a vector or data frame. 

```{r}
tab <- read.table("Data/airquality_output.txt", sep = ";", dec = ".", header = TRUE)
airquality$TempClass <- rep("", nrow(airquality))
airquality[airquality$Temp >= 70 , "TempClass"] <- "warm"
airquality[airquality$Temp < 70 , "TempClass"] <- "cold"
airquality$TempClass <- factor(airquality$TempClass)
```

## Flow control: *conditions*

With the `if()` statement you can also control which statements are executed, and sometimes this is more convenient. 


```{r}
i <- 55

if (i < 50) {
  print("Less than 50")
} else {
  print("Greater than 50")
}
```


## Flow control: *conditions*

You can combine multiple logical operations:

```{r}
i <- 55

if (i < 50) {
  print("Kleiner 50")
} else if ( i >= 50 & i < 70){
  print("Greater than 50 and less than 70")
} else {
  print("Greater than 70")
}
```

## Loops

The `for()` statement allows you to specify that a certain operation should be repeated a fixed number of times.

```{r}
x <- c(5, 4, 8)
for (i in x) {
  print(i^2)
}
```

The above example iteratively sets the variable `i` equal to each of the elements of vector `x`. For example, in the first loop iteration `i=5`, in the second iteration `i=4`, in the third and last iteration `i=8`. For each iteration, whatever commands are listed within the curly braces will be performed. The curly braces serve to group the commands so that they are treated by R as a single command.

## Loops

You can collect the results of the loop in a new variable. Based on the previous example, you may append the result of each iteration to a new vector `x_quadrat`.


```{r}
x <- c(5, 4, 8)

x_quadrat <- c() # create an empty vector to hold the results

for(i in x){
  #append the result to x_quadrat and store the new vector in x_quadrat
  x_quadrat <- c(x_quadrat, i^2)
}
x_quadrat
```

## Loops

Often it is more efficient and convinient to loop through an index vector rather than through a data vector. This way you can use indices to extract and store information from multiple variables (e.g. data frame columns). In the example below, `i` will loop through the vector c(1, 2, 3).

```{r}
x <- c(5, 4, 8)
y <- c(7, 2, 1)

xy_product = rep(NA, length(y))

for(i in 1:length(x)){
  xy_product[i] <- x[i] + y[i]
}

xy_product
```

## Loops

You can combine conditions and loops:

```{r}
# 100 random numbers between 1 and 100 sampled from a uniform distribution
random <- runif(100, min=1, max=100) 

count <- 0

for (i in random) {
  if (i < 50) {
    count <- count + 1
  }
}

count
```

## Loops

The `while()` function loops through an operation as long as a specified condition is true.

```{r}
i <- 0

while (i < 50) {
  i <- i + 10
  print(i)
}
```

## Loops vs vectorized operations

Can we conduct these vectorized operations with loops and conditions? 

```{r}
airquality$TempClass <- rep("", nrow(airquality))
airquality[airquality$Temp >= 70 , "TempClass"] <- "warm"
airquality[airquality$Temp < 70 , "TempClass"] <- "cold"
airquality$TempClass <- factor(airquality$TempClass)
```

Yes, we could achieve the same result by looping through rows with `for()` and then assigning classes using conditional statments with `if()`. However, loops are usully less efficient than vectorized operations in R!

## Functions

Functions are self-contained units with a well-defined purpose. In general, functions take inputs, do calculations (possibly printing intermediate results, drawing graphs, calling other functions, etc.), and produce outputs. The best thing is: You can write your own!

```{r}
my_addition_function <- function(x, y){
  z <- x + y
  return(z)
}

my_addition_function(x = 2, y = 4)
```

When the function is completed, all variables declared inside the function are gone. For example, the variable `z` is not known outside the scope of the function. To pass back results you can use `return()`.

## Functions

You can specify default values for variables as follows:

```{r}
my_addition_function <- function(x , y = 5){
  z <- x + y
  return(z)
}

my_addition_function(x = 2)
```

Those defaults are overwritten, if you pass it another value:

```{r}
my_addition_function(x = 2, y = 8)
```

## Functions

Functions can include loops and conditions:

```{r}
my_addition_function <- function(x, y, calculate = TRUE){
  if(calculate == TRUE){
    return(x + y)  
  }else if(calculate == FALSE){
    print("Nothing to calculate.")
  }
}

my_addition_function(x = 2, y = 3)

my_addition_function(x = 2, y = 3, calculate = FALSE)
```

## Functions

Usefull functions that are already integrated in R:

- `mean()`: Mittelwert
- `sum()`: Summe
- `var()`: Varianz
- `sd()`: Standardabweichung
- `median()`: Median
- `quantile()`: Quantile
- ...

## Functions

Lets write our own function for calculating the mean:

```{r}
mean(airquality$Temp)

my_mean <- function(x){
  mw <- sum(x, na.rm = TRUE) / length(x)
  return(mw)
}

my_mean(airquality$Temp)

```

## Functions

Note, the `mean()` function provides an argument to ignore missing values. Ours does not!

```{r}
mean(airquality$Ozone, na.rm = TRUE)
my_mean(airquality$Ozone)
```

## Functions

But we can fix that:

```{r}
mean(airquality$Ozone, na.rm = TRUE)

my_mean <- function(x){
  mw <- sum(x, na.rm = TRUE) / sum(!is.na(airquality$Ozone))
  return(mw)
}

my_mean(airquality$Ozone)

```

## Variable scope

The "scope" of a variable tells us where the variable would be recognized. We already discussed, variables inside functions are not known outside the function, i.e. they have a local scope (unless you declare them as global)!

HOWEVER: Variables defined at the console have global scope, meaning they are visible in any user-defined function. See variable `a` below. This can lead to unexpected results if you do not watch out.

```{r}
a = 10
my_addition_function <- function(x, y){
  z <- x + y + a
  return(z)
}

my_addition_function(x=2, y=4)
```

## Applying functions to matrices and data frames

The `apply()` function applies function to all rows or columns of a matrix or data frame:

```{r}
m <- matrix(1:9, nrow = 3)

apply(m, MARGIN = 1, FUN = mean)
apply(m, MARGIN = 2, FUN = mean)
```

The function must return a (single) scalar value. The resulting vector has length equal to `nrow()` or `ncol()`

## Applying functions to matrices and data frames

The `lapply()` function applies a function to each element of a list:

```{r}
lapply(list(a = 1, b = 2, c = 3), FUN = function(x){x * 2})
```

The result is also a list. The `sapply()` is a more user-friendly version of `lapply()` that returns simplefied results.


## Examples: `apply()`

Calculate the mean growth for each year from the treering dataset:

```{r fig.show='hide', warning=FALSE, message=FALSE}
library(ggplot2)

treering <- read.csv("Data/treering.csv")

growth_mean <- apply(treering[, 1:56], MARGIN = 1, FUN = mean, na.rm = TRUE)

ggplot(data = data.frame(Year = treering$Year, 
                         GrowthMean = growth_mean), 
       aes(x = Year, y = GrowthMean)) +
       geom_point()
```

## Examples: `apply()`

```{r echo=FALSE, results='hide', message=FALSE, warning=FALSE}
ggplot(data = data.frame(Year = treering$Year, 
                         GrowthMean = growth_mean), 
       aes(x = Year, y = GrowthMean)) +
  geom_point()
```

## Examples: `apply()`

We can also estimate and plot the 95% confidence interval:

```{r fig.show='hide', warning=FALSE, message=FALSE}
se <- function(x){sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))}
growth_se <- apply(treering[, 1:56], MARGIN = 1, FUN = se)

ggplot(data = data.frame(Year = treering$Year, 
                         GrowthMean = growth_mean,
                         SE = growth_se), 
       aes(x = Year, y = GrowthMean)) +
  geom_point() +
  geom_errorbar(aes(ymin = GrowthMean - 1.96 * SE, ymax = GrowthMean + 1.96 * SE))
```

## Examples: `apply()`

```{r echo=FALSE, results='hide', message=FALSE, warning=FALSE}
ggplot(data = data.frame(Year = treering$Year, 
                         GrowthMean = growth_mean,
                         SE = growth_se), 
       aes(x = Year, y = GrowthMean)) +
  geom_point() +
  geom_errorbar(aes(ymin = GrowthMean - 1.96 * SE, ymax = GrowthMean + 1.96 * SE))
```

## Die `tapply()` Funktion

The `tapply()` funktion applies a function to each element of a vector grouped by a second vector:

```{r}
tapply(1:9, INDEX = c(1, 1, 1, 2, 2, 2, 3, 3, 3), FUN = mean)
```

## Examples: `tapply()`

We can use `tapply()` to calculate mean and standard deviation of ozone concentration for each month:

```{r}
tapply(airquality$Ozone, INDEX = airquality$Month, FUN = mean, na.rm = TRUE)
tapply(airquality$Ozone, INDEX = airquality$Month, FUN = sd, na.rm = TRUE)
```
