---
title: "Assignment 14"
author: ""
date: ""
output: html_document
---

**Relationship between Picea Abies climate**

1) Calculate the count of Abies alba trees within each Bioclim grid cell (3 Points): 

Load the abies alba data and the bioclim data set used in Week 12. Create a new raster counting the number of Abies alba occurences in each grid cell.

```{r}
library(sp)
library(raster)
library(mapview)

abies <- read.csv("/Users/corneliussenf/Dropbox/Teching/Quantitative Methods/Week-14/data/abiesalba.csv")

abies_sp <- SpatialPoints(abies, CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
abies_sp <- spTransform(abies_sp, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"))

bioclim <- stack("/Users/corneliussenf/Dropbox/Teching/Quantitative Methods/Week-14/data/bioclim.tif")

countgrid <- rasterize(abies_sp, subset(bioclim, 1), fun = "count", background = 0)

mapview(countgrid)
```

2) Create a data frame from the data (3 Points): 

Stack the bioclim and count raster. Convert the stack into a data frame where each row equals one pixel. Remove NAs. Finally, create a stratified sample containing 10% of the observations in each count class. Hint: You can combine dplyr::group_by() and dplyr::sample_frac()!

```{r}
stck <- stack(countgrid, bioclim)
dat <- as.data.frame(stck)
names(dat) <- c("Count", paste0("Bioclim", 1:19))
dat_sample <- dat %>%
  na.omit() %>%
  filter(Count != 0) %>%
  group_by(Count) %>%
  sample_frac(0.1)
```

3) Build a model predicting the count of Picea Abies (3 Points)

Use a Poisson model for predicting the occurence of picea abies from bioclom. Either hand-select a variable combination or run a PCA model.

```{r}
ggplot(gather(dat_sample, key = bioclim, value = value, -Count), 
       aes(x = log(Count), y = value)) +
  geom_point() +
  facet_wrap(~bioclim, scales = "free")

cor(dat_sample[,-1])

gather(dat_sample, key = bioclim, value = value, -Count) %>%
  group_by(bioclim) %>%
  summarize(cor = cor(Count, value, method = "spearman")) %>%
  arrange(cor)

fit <- glm(Count ~ Bioclim10 * Bioclim14, data = dat_sample, family = poisson("log"))

summary(fit)
```

4) Create a map of Picea abies counts (1 Point)

```{r}
names(bioclim) <- paste0("Bioclim", 1:19)

prediction <- predict(bioclim, fit)

mapview(prediction)
```

