---
title: "Spatial statistics in R"
author: "Cornelius Senf"
date: "cornelius.senf@geo.hu-berlin.de"
output:
  ioslides_presentation:
    smaller: yes
    widescreen: yes
  beamer_presentation: default
  slidy_presentation: default
subtitle: Quantitative Methods - Lab Session 14
header-includes: \usepackage{amsmath}
---

<style type="text/css">
slide.backdrop {
  background: none !important;
  background-color: white !important;
}
</style>

```{r echo=FALSE, hide=TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
library(tidyverse)
```

## Today

- Descriptive point pattern analysis

## Data

We are going to use the tree species dataset from two weeks ago, more specifically the occurences of Abies alba (silver fir):

```{r, message=FALSE, warning=FALSE}
library(sp)
library(raster)

abies <- read.csv("data/abiesalba.csv")

head(abies)

abies_sp <- SpatialPoints(abies, CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))

abies_sp <- spTransform(abies_sp, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"))
```

## Data

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(sp)
library(raster)
library(mapview)

abies <- read.csv("data/abiesalba.csv")
abies_sp <- SpatialPoints(abies, CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
abies_sp <- spTransform(abies_sp, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"))
mapview(abies_sp)
```

# Descriptive point pattern

## Descriptive point pattern

The centroid can easily be calculated by averaging the coordinates:

```{r}
coords <- abies_sp@coords
coords <- as.data.frame(coords)

xmean <- mean(coords[, 1])
ymean <- mean(coords[, 2])

p <- ggplot(coords, aes(x = long, y = lat)) + 
  geom_point(alpha = 0.3) +
  geom_point(aes(x = xmean, y = ymean), col = "red", shape = 2)
```

## Descriptive point pattern

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center'}
print(p)
```

## Descriptive point pattern

We can also calculate the median of the coordinates:

```{r}
xmedian <- median(coords[, 1])
ymedian <- median(coords[, 2])

p <- ggplot(coords, aes(x = long, y = lat)) + 
  geom_point(alpha = 0.3) +
  geom_point(aes(x = xmean, y = ymean), col = "red", shape = 2) +
  geom_point(aes(x = xmedian, y = ymedian), col = "blue", shape = 2)
```

## Descriptive point pattern

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center'}
print(p)
```

## Descriptive point pattern

We can calculate the average/minimum/maximum distance between points:

```{r}
dist <- raster::pointDistance(abies_sp, longlat = FALSE)
dist[1:3, 1:3]

dist <- dist[lower.tri(dist)]

c(min(dist), mean(dist), max(dist)) / 10000
```

## Point counts

We can generate a grid from the points extent::

```{r}
ext <- extent(abies_sp)
ext

ras <- raster(ext)
ras

```

## Point counts

And set the resolution to 1000*1000 meters while keeping the projection:

```{r}
dim(ras) <- c(1000, 1000, 1)
ras

projection(ras) <- projection(abies_sp)
```

## Point counts

Following we can use this raster to rasterize the points and count the number of points in each cell:

```{r, eval=FALSE}
countgrid <- rasterize(abies_sp, ras, fun = "count", background = 0)

mapview(countgrid)
```

## Point counts

```{r, echo=FALSE, warning=FALSE, message=FALSE}
countgrid <- rasterize(abies_sp, ras, fun = "count", background = 0)

mapview(countgrid)
```

## Distances as raster

We can easily calculate the distances to cells with Abies. However, I first aggregate the raster to a loer resolution (otherwise it will take too long):

```{r, eval=FALSE}
abies_ras <- aggregate(countgrid, fact = 10, fun = sum)

mapview(abies_ras)
```

## Distances as raster

```{r, echo=FALSE, warning=FALSE, message=FALSE}
abies_ras <- aggregate(countgrid, fact = 10, fun = sum)

mapview(abies_ras)
```

## Distances as raster

Following I set all cells containing no Abies to NA and calculate the distance:

```{r, eval=FALSE}
abies_ras[abies_ras == 0] <- NA

distance_abies <- distance(abies_ras)

mapview(distance_abies)
```

## Distances as raster

```{r, echo=FALSE, warning=FALSE, message=FALSE}
abies_ras[abies_ras == 0] <- NA

distance_abies <- distance(abies_ras)

mapview(distance_abies)
```

## Kernel estimates

Kernel density estimates do not simply count the occurences, but 'smooth' by fitting a function (e.g., multivariate-normal) to the data. We can calculate Kernel densities as raster:

```{r, eval=FALSE, warning=FALSE, message=FALSE}
library(spatialEco)

kernel_ras <- sp.kde(abies_sp, bw = 10000, newdata = ras)

mapview(kernel_ras)
```

## Kernel estimates

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
library(spatialEco)

kernel_ras <- sp.kde(abies_sp, bw = 10000, newdata = ras)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
mapview(kernel_ras)
```

## Kernel estimates

ggplot2 can also estimate kernels and nicely plot them atop of Google Maps!

```{r, warning=FALSE, message=TRUE}
library(ggmap)

map <- get_map(location = c(lon = mean(abies$long), lat = mean(abies$lat)), zoom = 5)

p <- ggmap(map) + 
  geom_density2d(data = abies, aes(x = long, y = lat), size = 0.3) + 
  stat_density2d(data = abies, 
                 aes(x = long, y = lat, fill = ..level.., alpha = ..level..), size = 0.01, 
                 bins = 50, geom = "polygon") + 
  scale_fill_gradient(low = "green", high = "red") + 
  scale_alpha(range = c(0.1, 0.5), guide = FALSE)
```

## Kernel estimates

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=6, fig.align='center'}
print(p)
```
