---
title: "Assignment 1"
author: "Cornelius Senf"
date: "17. October 2017"
output: html_document
---

### 1. Create a vector from a series of numbers:

2-2-2-2-2-2-2-2-2-2-2-8-8-8-8-1-2-3-4-5-6-7-8-9-10-6-5-4-3-2-1-56-56-56-56-8-8-8-8-8

From the above number series construct a vector `x` using the functions c(), rep() und seq(). Note, the elements in the above number series are separated by minus signs for readability!

```{r}
x <- c(rep(2, 11),
         rep(8, 4),
         1:10,
         seq(from = 6, to = 1, by = -1),
         rep(56, 4),
         rep(8, 5))
```

### 2. How many elements has vector x?

```{r}
length(x)
```

### 3. Extract the 12th element from vector x.

```{r}
x[12]
```

### 4. Extract the elements 20-27 from vector x.

```{r}
x[20:27]
```

### 5. Extract the 12th and 20-27th element from vector x.

```{r}
x[c(12, 20:27)]
```

### 6. Convert vector x into a matrix m with 5 rows und 8 columns. Show the outputs of matrix m.

```{r}
m <- matrix(x, nrow = 5)
m
```

### 7. Extract the third column from matrix m.

```{r}
m[, 3]
```

### 8. From matrix m, extract the element from the fourth row and second column,

```{r}
m[4, 2]
```

### 9. From matrix m, extract the 7ths and 8ths column.

```{r}
m[, 7:8]
```
