---
title: "Assignment 1"
author: "Name"
date: "Date"
output: html_document
---

### 1. Create a vector from a series of numbers:

2-2-2-2-2-2-2-2-2-2-2-8-8-8-8-1-2-3-4-5-6-7-8-9-10-6-5-4-3-2-1-56-56-56-56-8-8-8-8-8

From the above number series construct a vector `x` using the functions c(), rep() und seq(). Note, the elements in the above number series are separated by minus signs for readability!

```{r}

```

1 Point

### 2. How many elements has vector x?

```{r}

```

1 Point

### 3. Extract the 12th element from vector x.

```{r}

```

1 Point

### 4. Extract the elements 20-27 from vector x.

```{r}

```

1 Point

### 5. Extract the 12th and 20-27th element from vector x.

```{r}

```

1 Point

### 6. Convert vector x into a matrix m with 5 rows und 8 columns.

```{r}

```

1 Point

### 7. Extract the third column from matrix m.

```{r}

```

1 Point

### 8. From matrix m, extract the element from the fourth row and second column,

```{r}

```

1 Point

### 9. From matrix m, extract the 7ths and 8ths column.

```{r}

```

1 Point

### 10. Convert matrix m into a data.frame names 'mm' and extract the 3rd column

```{r}

```

1 Point
