---
title: "Introduction to R"
author: "Cornelius Senf"
date: "cornelius.senf@geo.hu-berlin.de"
output:
  ioslides_presentation:
    smaller: yes
    widescreen: yes
  beamer_presentation: default
  slidy_presentation: default
subtitle: Quantitative Methods - Lab Session 11
header-includes: \usepackage{amsmath}
---

<style type="text/css">
slide.backdrop {
  background: none !important;
  background-color: white !important;
}
</style>

```{r echo=FALSE, hide=TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
library(tidyverse)
```

## The 4 R Sessions:

1. Basic concept of programming, data types, vectors and factors

2. Matrices and data.frames, data import/export, graphics and visualization

3. Tidy data, data wrangling

4. **Today:** Spatial data in R 

# Raster data in R

## Raster data in R

```{r message=FALSE, warning=FALSE}
library(raster)
```

## Raster data in R

```{r}
ras <- raster(ncol = 100, nrow = 100, xmn = 0, xmx = 100, ymn = 0, ymx = 100)

ras
```

## Raster data in R

```{r fig.show='hide'}
m <- rnorm(n = ncell(ras), mean = 0, sd = 1)

values(ras) <- m

plot(ras)
```

## Raster data in R

```{r, echo=FALSE, results='asis', warning=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center'}
plot(ras)
```

## Raster data in R

```{r}
ras2 <- ras + 10

ras3 <- ras * ras2

ras4 <- ras^2

ras5 <- (ras - cellStats(ras, stat = mean)) / 
                                cellStats(ras, stat = sd)
```

## Raster data in R

```{r, echo=FALSE, results='asis', warning=FALSE, message=FALSE, fig.height=5.5, fig.width=5.5, fig.align='center'}
par(mfrow=c(2, 2), mar=c(2.5, 2.5, 3, 3))
plot(ras2, main = "ras + 10")
plot(ras3, main = "ras * ras2")
plot(ras4, main = "ras^2")
plot(ras5, main = "z(ras)")
```

## Raster data in R

```{r}
ras.na <- ras

ras.na[ras.na > 0.5] <- NA
```

## Raster data in R

```{r, echo=FALSE, results='asis', warning=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center'}
plot(ras.na)
```

## Raster data in R

```{r}
k <- matrix(c(  0, 0.5, 1,
              0.5, 1.5, 2,
              1.5, 6.0, 3), 
            ncol = 3, byrow = TRUE)
k

ras.reclass <- reclassify(x = ras, rcl = k)
```

## Raster data in R

```{r, echo=FALSE, results='asis', warning=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center'}
plot(ras.reclass, col = c("black", "blue", "red", "green"))
```

## Raster data in R

```{r}
ras[10, 20]

ras[10, 20] <- 10
```

## Raster data in R

```{r, echo=FALSE, results='asis', warning=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center'}
plot(ras)
```

## Raster data in R

```{r}
ras.stack <- stack(ras, ras2)

ras.stack
```

## Raster data in R

```{r}
layer1 <- subset(x = ras.stack, subset = 1)

layer1
```

## Raster data in R

```{r}
data <- as.data.frame(ras.stack)

head(data)
```

## Raster data in R

```{r fig.show='hide', message=FALSE}
elevation <- raster("Data/elevation.envi")

elevation
```

## Raster data in R

```{r}
projection(elevation)
```

See for projection definitions: http://spatialreference.org/

## Raster data in R

```{r fig.show='hide', message=FALSE}
plot(elevation)
```

## Raster data in R

```{r, echo=FALSE, results='asis', warning=FALSE, message=FALSE, fig.height=5, fig.width=6, fig.align='center'}
plot(elevation)
```

## Raster data in R

```{r}
climate <- stack("Data/climate.envi")

nlayers(climate)

names(climate)

names(climate) <- c("Tmax", "Tmin", "Tmean", "Prec")
```

## Export raster data

```{r}
writeRaster(elevation, 
            filename= "Data/elev.envi",
            overwrite = TRUE)
```

# Vector data

## Vector data

```{r, message=TRUE, warning=TRUE}
library(sp)
```

## Vector data: simple random sample

```{r}
coords_srs <- data.frame(x = runif(100, 17, 27), 
                         y = runif(100, 45, 50))

pts_srs <- SpatialPoints(coords = coords_srs)
pts_srs
```

## Vector data: Assign Coordinate system

```{r fig.show='hide'}
proj <- projection(elevation)

CRS(proj)

pts_srs <- SpatialPoints(coords = coords_srs,
                    proj4string = CRS(proj))
```

## Vector data

```{r, echo=TRUE, results='asis', warning=FALSE, message=FALSE, fig.height=5, fig.width=6, fig.align='center'}
plot(elevation)
plot(pts_srs, add = TRUE)
```

## Advanced plotting

```{r, eval=FALSE, message=FALSE, warning=FALSE}
library(mapview)

mapview(climate)
```

## Advanced plotting

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(mapview)
mapview(climate)
```

## Vector data

```{r}
df.climate <- extract(climate, pts_srs, df = TRUE)

class(df.climate)
head(df.climate)
```

## Vector data

```{r}
spdf.climate <- extract(climate, pts_srs, sp = TRUE)

spdf.climate
```

## Vector data

You can acess SpatialPointsDataFrame the same way as data.frame!

```{r}
head(spdf.climate)

spdf.climate$Tmax[1]
```

## Reading vector data

```{r, message=FALSE, warning=FALSE}
library(rgdal)
```

## Reading vector data

```{r, message=FALSE, warning=FALSE}
pts.sample <- readOGR('Data', 'pts_sample')
pts.sample
```

## Writing vector data

```{r, message=FALSE, warning=FALSE}
writeOGR(spdf.climate, "Data", "pts_climate", driver = "ESRI Shapefile", overwrite_layer = TRUE)
```
