---
title: "Assignment 5"
author: ""
date: ""
output: html_document
---

### 1. Fit a model explaining Ozone concentration by Temperature (2 Points)

Use the `lm()` function to fit a model explaining ozone concentration by temperature. Use the model to answer the following questions: (1) How does Ozone concentration change with Temperature? (2) Is the effect of temperature on ozone concentration significant? (3) How much of the varianze in ozone concentration is explained by temperature alone?

```{r}

```

### 4. Plot the regression (2 Points)

Create a plot showing the relationship between temperature and ozone concentration. Also show uncertainties caused by uncertainty in the parameter estimates (confidence intervall).

```{r}

```

### 3. Residual plot analysis (3 Points)

Create two residual plots for the model: a qq-plot and a scatterplot of fitted values versus residuals. Do the residuals look normal and properly distributed? Answer the question in one to three sentences.

```{r}

```

### 4. Transforming the response (3 Points in total, each part 1 point)

4.1 Create a new column 'logOzone' in the airquality dataset by calculating the natural logarithm of Ozone concentration. Following, fit a model explaining the logarithem of ozone concentration by temperature.

```{r}

```

4.2 Check the residuals of the model. Do they look different/better/worse than for the previous model?

```{r}

```

4.3 Plot the regression line of the logged model (without confidence intervals). You can either plot it in log-space (using the logged ozone concentrations) or back-transform the logged ozone concentrations into linear space using the exp() function (recall: exp(ln(x)) = x).

```{r}

```

