---
title: "Assignment Lab Session 2"
author: "Cornelius Senf"
date: "18.10.2017"
output: html_document
---

### 1. Vectors (1 Point)

Create a vector from a series of numbers:

2-2-2-2-2-2-2-2-2-2-2-8-8-8-8-1-2-3-4-5-6-7-8-9-10-6-5-4-3-2-1-56-56-56-56-8-8-8-8-8

From the above number series construct a vector `x` using the functions c(), rep() und seq(). Note, the elements in the above number series are separated by minus signs for readability!

```{r}
x <- c(rep(2, 11),
         rep(8, 4),
         1:10,
         seq(from = 6, to = 1, by = -1),
         rep(56, 4),
         rep(8, 5))
```

How many elements has vector x?

```{r}
length(x)
```

Extract the 12th element from vector x.

```{r}
x[12]
```

Extract the elements 20-27 from vector x.

```{r}
x[20:27]
```

Extract the 12th and 20-27th element from vector x.

```{r}
x[c(12, 20:27)]
```

### 2. Matrices (1 Point)

Convert vector x into a matrix m with 5 rows und 8 columns.

```{r}
m <- matrix(x, nrow = 5)
m
```

Extract the third column from matrix m.

```{r}
m[, 3]
```

From matrix m, extract the element from the fourth row and second column,

```{r}
m[4, 2]
```

From matrix m, extract the 7ths and 8ths column.

```{r}
m[, 7:8]
```

### 3. Data frames (2 Points)

Download the airquality data set from Moodle and import it into R.

```{r}
dat <- read.csv("Data/airquality.txt")
```

Answer the following questions:

- What is the concentration of ozone on May 20th?
- What is the mean temperature on July 16th?
- How many days had an ozone concentration greater than 50 ppm?

```{r}
# What is the concentration of ozone on May 20th?
dat[dat$Month == 5 & dat$Day == 20, "Ozone"]

# What is the mean temperature on July 16th?
dat[dat$Month == 7 & dat$Day == 16, "Temp"]

# How many days had an ozone concentration greater than 50 ppm?
sum(dat$Ozone > 50, na.rm = TRUE)
```

### 4. Manipulate data (3 Points)

Convert column `Month` (numeric) to a factor using the months names as factor labels ("May", "June", ...).

```{r}
dat$Month <- factor(dat$Month)
levels(dat$Month) <- c("May", "June", "July", "August", "September")
```

Use the function `paste()` to create a new column called 'Date' of the following format: 'Month-Day', 'May-1', 'May-2', 'May-3', ...

```{r}
Date <- paste(dat$Month, dat$Day, sep = "-")
dat <- cbind(dat, Date)
```

Create two new variables ('Temp.C' and 'Wind.ms') that show temperature in Celsius and wind speed in m/s (instead of Fahrenheit und mph). Show the first five records for each new variable.

```{r}
# Temperature
dat$Temp.C <- (dat$Temp - 32) / 1.8
head(dat$Temp.C)

# Windpseed
dat$Wind.ms <- dat$Wind * 0.44704
head(dat$Wind.ms)
```

Write the data to a file called 'airquality2.csv' using scv format.

```{r}
write.csv(dat, file = "Data/airquality2.csv", row.names = FALSE)
```

### 5. Loading new data (3 points)

Download the dataset 'treering.csv' from Moodle. Inspect it in either Excel or a text editor (Notepad++ for Windows or TextWrangler for Mac). Try loading the dataset into a data frame.

The dataset contains the width of treerings measures at several plots.

```{r}
treering <- read.table("Data/treering.csv", sep = ";", dec = ",", header = TRUE)
```

Drop all rows containing at least one `NA` value.

```{r}
treering_nomissing <- na.omit(treering)
```

What is the mean tree-ring width for plot X40B?

```{r}
mean(treering_nomissing$X40B)
```

