---
title: "Assignment Lab Session 2"
author: "Name"
date: "Date"
output: html_document
---

### Steps to complete your assignment:

1. Complete the tasks. Write your R code in the designated areas (you can send the code to the console to test your code).
2. Create an HTML file in R-Studio: | Menu | File | Knit Document (alternatively: little Knit button in script header)
3. RENAME the file: lastname_assignment01.html
4. Upload your assignment (.html file) to Moodle

### 1. Vectors (1 Point)

Create a vector from a series of numbers:

2-2-2-2-2-2-2-2-2-2-2-8-8-8-8-1-2-3-4-5-6-7-8-9-10-6-5-4-3-2-1-56-56-56-56-8-8-8-8-8

From the above number series construct a vector `x` using the functions c(), rep() und seq(). Note, the elements in the above number series are separated by minus signs for readability!

```{r}

```

How many elements has vector x?

```{r}

```

Extract the 12th element from vector x.

```{r}

```

Extract the elements 20-27 from vector x.

```{r}

```

Extract the 12th and 20-27th element from vector x.

```{r}

```

### 2. Matrices (1 Point)

Convert vector x into a matrix m with 5 rows und 8 columns.

```{r}

```

Extract the third column from matrix m.

```{r}

```

From matrix m, extract the element from the fourth row and second column.

```{r}

```

From matrix m, extract the 7ths and 8ths column.

```{r}

```

Convert matrix m into a data.frame names 'mm' and extract the 3rd column

```{r}

```

### 3. Data frames (2 Points)

Download the airquality data set from Moodle and import it into R.

```{r}

```

Answer the following questions:

- What is the concentration of ozone on May 20th?
- What is the mean temperature on July 16th?
- How many days had an ozone concentration greater than 50 ppm?

```{r}

```

### Manipulate data (3 Points)

Convert column `Month` (numeric) to a factor using the months names as factor labels ("May", "June", ...).

```{r}

```

Use the function `paste()` to create a new column called 'Date' of the following format: 'Month-Day', 'May-1', 'May-2', 'May-3', ...

```{r}

```

Create two new variables ('Temp.C' and 'Wind.ms') that show temperature in Celsius and wind speed in m/s (instead of Fahrenheit und mph). Show the first five records for each new variable.

```{r}

```

Write the data to a file called 'airquality2.csv' using scv format.

```{r}

```

### Loading new data (3 points)

Download the dataset 'treering.csv' from Moodle. Inspect it in either Excel or a text editor (Notepad++ for Windows or TextWrangler for Mac). Try loading the dataset into a data frame.

The dataset contains the width of treerings measures at several plots.

```{r}

```

Drop all rows containing at least one `NA` value.

```{r}

```

What is the mean tree-ring width for plot X40B?

```{r}

```

