---
title: "Introduction to R"
author: "Cornelius Senf"
date: "cornelius.senf@geo.hu-berlin.de"
output:
  ioslides_presentation:
    smaller: yes
    widescreen: yes
  beamer_presentation: default
  slidy_presentation: default
subtitle: Quantitative Methods - Lab Session 02
header-includes: \usepackage{amsmath}
---

<style type="text/css">
slide.backdrop {
  background: none !important;
  background-color: white !important;
}
</style>

## Goals of today' session:

- Matrices and data frames
- Data import and export
- Simple data manipulation

# Matrices and data frames

## R classes

Recall the most important object classes in R:

- vector
- factor
- matrix
- data frame
- list

We already introduced you to vectors and factors. Today we will cover **matrix**, **data.frame** and **list** objects.

## Matrices

A matrix is a two-dimensional **Array**. A matrix therefore has additional **attributes** specifiying the 2 dimensions: **nrow** und **ncol**.

```{r}
m <- matrix(1:9, nrow = 3, ncol = 3)
m
```

See: `dim(m)`, `nrow(m)`, `ncol(m)`

## Matrices

By default the columns of the matrix will be filled first.  If you want to fill the matrix by row, you can specify this with the `byrow` argument:

```{r}
n <- matrix(1:9, nrow = 3, ncol = 3, byrow = TRUE)
n
```

## Matrices

You can apply mathematical operators to matrices the same way as vectors (Attention: Recycling rule!): 

```{r}
m * 2
m * n
```

## Matrices

Like with vectors, you can access elements of a matrix with indices, except that we now deal with two dimensions `[i,j]` or `[row, column]`:

```{r}
m
m[1, ]
m[ , 1]
```

## Matrices

```{r}
m[1, 1]
m[1:2, 3]
m[1:2, c(1,3)]
```

## Matrices

When you exctract elements from a matrix, the result can belong to a different class!

```{r}
class(m)
class(m[ , 3])
```

## Matrices

The functions `cbind()` und `rbind()` glue (bind) vectors and matrices together:

```{r}
# bind together by column
cbind(m,n)
```

## Matrices

The function `cbind` und `rbind` glue (bind) vectors and matrices together:

```{r}
# bind together by row
rbind(m,n)
```

## Lists

Lists are objectes that can store values of different data type (and different objects):

```{r}
l <- list(c(1, 2, 3), m, "a")
```

To access the elements of a list with indices you need to use double brackets `[[]]`:

```{r}
l[[1]]
l[[3]]
```

## Why bother with lists now?

Nearly all functions (e.g., linear regression, generalised linear modelling, t-test, etc.) in R produce output that is stored in a list. The good news is that R contains various functions to extract required information (e.g., estimated values, p-values, etc.) and presents it in nice tables. However, sometimes it can be useful to extract information from the results directly OR you may want to use a list to return the output from your own functions.

## Lists

```{r}
class(l)
class(l[2])
class(l[[2]])
```

## Data frames

A data frame is what you may call a data table. It is similar to a two-dimensional matrix but the columns can contain different data types.

```{r}
df <- data.frame(TREEID = 1001:1003, 
                 SPECIES = factor(c("Spruce", "Fir", "Pine")), 
                 LIFE = c(TRUE, FALSE, TRUE),
                 HEIGHT = c(34, 21, 26)
                 )
df
```

## Data frames

The `summary()` function gives a quick overview. Helpful for spotting data entry errors and `NA`'s:

```{r}
summary(df)
```

## Data frames

You can index (access) columns using three main ways:

```{r}
df$TREEID
df[ , 1]
df[ , "TREEID"]
```

## Data frames

Rows are indexed by row number:

```{r}
df[3, ]
df[1:2, "TREEID"]
df[1, c("TREEID", "HEIGHT")]
```

## Data frames

IMPORTANT: Extracting a row does not change the class but, extracting a column does!

```{r}
class(df)
class(df[ 1, ])
class(df[ , "TREEID"])
```

# Data import and export

## Data import

R can read a variety of dataset formats:

- Text files (e.g. CSV, TXT)
- Statistical programs (e.g. Excel, SPSS table)
- DBF file (e.g. ArcGIS)
- Databases (e.g. PostgreSQL)
- local file system or on a remote server (e.g. ftp, http)
- https://cran.r-project.org/doc/manuals/r-release/R-data.html

## Data import: Text files

Text (ASCII) files are a popular data storage and exchange format. They can be read on any OS platform and without specialty software (Notepad).

Comma-separated (csv) or tab-separated files are common formats to store table data (in systems with English locale). 

BE AWARE: On German (and other) systems, the comma is already reserved for decimal places, so here the semi-colon or tab-separation is sometimes prefered. On English systems, the dot (`.`) is used for decimal places, on other systems it may be used to group digits for readability, e.g. `1,000,000`.

## Data import: Text files

The `read.table()` function can be used to read table data from text files. The function allows several arguments to accomodate the different data formats. See `?read.table()`

The following example specifies the following format options:

- The columns are separated by `,`
- The decimal sign is `.`
- and the first row contains the column names `header = TRUE`

```{r}
tab <- read.table("Data/airquality.txt", sep = ",", dec = ".", header = TRUE)
```

## Data import: Text files

`read.table()` returns a data frame.

```{r}
class(tab)
names(tab)
head(tab, 3)
```

## Data import: Text files

`read.csv()` is a short-cut of `read.table()` text files with comma-separated columns.

## Data export: Text files
Export data frames to semi-colon delimited file:

```{r}
write.table(tab, "Data/airquality_output.txt", sep = ";", dec = ".", row.names = FALSE)
```

You can also use `write.csv()` to export data frames to comma-delimited text files:

```{r}
write.csv(tab, "Data/airquality_output.csv")
```

    write.table(x, file = "", append = FALSE, quote = TRUE, sep = " ",
    eol = "\n", na = "NA", dec = ".", row.names = TRUE,
    col.names = TRUE, qmethod = c("escape", "double"),
    fileEncoding = "")
            
# Simple data manipulation

## Missing values

Recall from last session that `NA` is used for missing values in R: 

```{r}
x <- c(1, 5, 3, 6, NA, 9, 21, 4)
x
```

..and that you must use `is.na()` to determine if an element is or contains missing values.

```{r}
is.na(x)
```

## Missing values

It is not uncommon to have missing values in datasets, e.g. in data frames and matrices:

```{r}
df <- data.frame(var1 = c(1, 3, 12, NA, 5), 
                 var2 = c(3, 4, 1, 8, 11))
df
```

## Missing values

Use `na.omit()` to ignore rows in a data frame that contain `NA`s:

```{r}
na.omit(df)
```

## Missing values

Also recall that arithmetic functions and operations applied to `NA`s return `NA`

```{r}
NA * 3
```

Many arithmetic functions allow you to specify whether to ignore or include `NA`s:

```{r}
sum(df$var1)
sum(df$var1, na.rm=TRUE)
```

## Text manipulation

Combine two or more character variables witht `paste()`:

```{r}
paste("Hello", "World", sep = "_")
```

Extract a portion of a character variable `substring()`:

```{r}
substring("Hello World", first = 3, last = 8)
```

## Subset columns of a data frame

It is best to extract columns from a data frame using the column names:

```{r}
names(airquality)

airquality_ozone <- airquality$Ozone
airquality_ozone <- airquality[, "Ozone"]
airquality_ozone_temp <- airquality[, c("Ozone", "Temp")]
```

## Subset rows of a data frame

Subset rows based using logical operations on variables (columns):

```{r}
airquality_temp_gr_70 <- airquality[airquality$Temp > 70, ]
nrow(airquality_temp_gr_70)
```

Subset rows based on row indices:
```{r}
airquality_zeile_10_100 <- airquality[1:100, ]
nrow(airquality_zeile_10_100)
```

## Subset rows of a data frame

You can combine logical operators to make more complex subsets of rows and columns:

```{r}
airquality_juni <- airquality[airquality$Month == 6, ]
```

all measurements from 15. Juni:

```{r}
airquality_juni_15 <- airquality[airquality$Month == 6 & airquality$Day == 15, ]
```

or return only Ozone values (column) from 15. Juni:

```{r}
ozone_juni_15 <- airquality[airquality$Month == 6 & airquality$Day == 15, "Ozone"]
```

## Subset rows of a data frame

In previously covered logical operations, single values were compared to vectors or matices (or single values), e.g. `c(1,2,3) < 2`.

The `%in%` operator can be applied to two vectors: `x %in% y`. For each element in vector `x`, `%in%` evaluates if the element is contained in vector `y`. The operator returns a logical vector of the same lenght as vector `x`.

The following example returns all rows of month June and July.
```{r}
airquality_jun_jul <- airquality[airquality$Month %in% c(6, 7), ]
```

## Create a new variable from existing variables

Recall, the `$` operator is used to access columns in a data frame by name. You can also use it to create a new column:

The following example creates a new column with the name "NewVariable" and fills it with Ozone values multiplied by 100.

```{r}
airquality$NewVariable <- airquality$Ozone * 100
```

Or we create a log transformated variable.

```{r}
airquality$logOzone <- log(airquality$Ozone)
```

## Create a factor variable from numeric variables

The following example creates a factor (categorical) variable from temperature data:

First, add an empty character vector:

```{r}
airquality$TempClass <- rep("", nrow(airquality))
```

Then, recode the character vector based on logical operations on `$Temp`:

```{r}
airquality[airquality$Temp >= 70 , "TempClass"] <- "warm"
airquality[airquality$Temp < 70 , "TempClass"] <- "cold"
airquality$TempClass <- factor(airquality$TempClass)
```

## Create a factor variable from numeric variables

The functino `cut` divides a numeric variable into intervals and codes them into factors (categorical data):

```{r}
airquality$TempClass2 <- cut(airquality$Temp, c(0, 69, 100), labels=c('cold', 'warm'))
```

Note, there are even more ways to do the same task!!
