---
title: "Assignment 7"
author: "Cornelius Senf"
date: "28.11.2017"
output: html_document
---

**Exercise:** In an experiment, the diameter of Tetrahymena cells (free-living ciliate protozoa) was measured in cultures of varying cell density and with and without glucose supplement (= additional energy). Using a linear regression approach, analyze the data to address the following questions: (1) Does cell concentration influence cell growth? (2) If so, does the effect vary between cell cultures with glucose supplement and witout?

## 1. Import and prepare the dataset (1 Point)

The data is on Moodle.

```{r}
dat <- read.table("Data/tetrahymena_zellen.txt", sep = "\t", header = TRUE)
dat$glucose <- factor(dat$glucose)
```

## 2. Visualize the data (2 Points)

Create a scatterplot to explore the relationships between concentration, diameter and glucose supplement. Is the relationship linear? If not, how can you correct for it? Hint: Transforming a predictor might help! Briefly describe your thoughts on the relationships and the effects you might expect.

```{r}
library(ggplot2)

ggplot(dat, aes(x = concentration, y = diameter, col = glucose)) +
  geom_point() +
  geom_smooth(method = "lm")
```

Log-transforming the concentration improves the fit as concentration is boundeed to positive values.

```{r}
ggplot(dat, aes(x = log(concentration), y = diameter, col = glucose)) +
  geom_point() +
  geom_smooth(method = "lm")
```

## 3. Build a simple linear model (2 Points)

Build an appropriate linear model between diameter (response) and concentration (predictor) and visualize it graphically. Assess the assumptions and the goodness-of-fit of the model. Briefly describe your findings.

```{r}
fit <- lm(diameter ~ log(concentration), data = dat)
summary(fit)
```

Compare fitted values to residuals:

```{r}
resids <- data.frame(residuals = residuals(fit),
                     fitted = fitted(fit))

ggplot(resids, aes(x = fitted, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0)
```

Check for normality of residuals:

```{r}
ggplot(resids, aes(sample = residuals)) +
  stat_qq() +
  geom_abline(intercept = 0, slope = 1)
```


## 4. Multiple linear model (2 Points)

Now build a multiple linear model by adding glucose as additional predictor interacting with concentration. Compare your model to the previous one based on the AIC. Does including glucose improve the model fit?

```{r}
fit2 <- lm(diameter ~ log(concentration) * glucose, data = dat)
summary(fit2)
```

Comparing both models using AIC:

```{r}
AIC(fit, fit2)
```

AIC suggests that fit2 should be preferred over fit (smaller AIC).

## 5. Plot the regression lines of the multiple linear model (3 Points)

Plot the regression line for the multiple-regression model, that is one line for each class (glucose/no glucose).

```{r, message=FALSE, warnings=FALSE}
library(effects)

efct <- effect("log(concentration):glucose", mod = fit2, 
               xlevels = list(concentration = seq(1e+04, 6e+05, length.out = 100)),
               confint = list(compute = TRUE))
efct <- as.data.frame(efct)

ggplot() +
  geom_ribbon(data = efct, aes(x = concentration, ymin = lower, ymax = upper, fill = glucose), alpha = 0.2) +
  geom_point(data = dat, aes(x = concentration, y = diameter, col = glucose)) +
  geom_line(data = efct, aes(x = concentration, y = fit, col = glucose)) +
  scale_color_brewer(palette = "Set1") + # See http://colorbrewer2.org/ for hints on nice color-schemes!
  scale_fill_brewer(palette = "Set1")
```


